Compiling Python Modules to Native Parallel Modules Using Pythran and OpenMP Annotations

### 2017-10-26 docs/papers/sc2013/paper.tex
Read The Docs integration

Another widely used solution is the module IPython.parallel, which supports many
different styles of parallelism including: single program, multiple data (SPMD)
parallelism, multiple program, multiple data (MPMD) parallelism. message passing
using MPI, task farming, and data parallel.
FIXME: to be continued...

@@

Figure~\ref{fig:pythran-compiler} summarizes its processing: type inference
helps removing most dynamic behavior, and is combined with a high-level runtime
library ---\texttt{Pythonic++}--- to allow a one-to-one mapping between Python
and C++ constructs. Pythonic++ is a template library, i.e.\ it consists
exclusively of a set of headers. Therefore the generated C++ file is easily
distribuable and does not depend on Pythran. One could even preprocess this
C++ file to avoid the need of distributing along the Pythonic++ headers.

@@

At the AST level, it means that the Python AST is first reduced to a tree where
all nodes not available in C++ are transformed into a compatible representation.
For instance, list comprehension expressions are turned into function calls or
tuple unpacking is turned in multiple assignments. During this transformation
process, an expression is always transformed into a single expression and a
statement is always transformed into a single statement.
% FIXME as noted by reviewer 4, the previous sentence is not exact. Just a few
% lines above we said that "tuple unpacking is turned in multiple assignments"
Then this AST is converted into a C++ AST meant to be pretty-printed.

@@

The traditional solution is to use a native type that exposes a Python interface
and has a constant translation cost. NumPy's \texttt{ndarray} is a typical
implementation of this concept and is the keystone of the Scipy and NumPy
packages. Basically, a NumPy \texttt{ndarray} is a raw C pointer that is exposed
at the Python level. The PEP 3118 defined the API that allows to share
efficiently the embeded data between Python and the native world without
involving any copy. As a tool for scientific computing, Pythran supports such a
structure, implemented as a class wrapping a raw pointer.

@@

The \texttt{reduction(\emph{operator}: \emph{list})} directive is used to
characterize some data dependencies when performing a parallel reduction. The
list of supported operators depend on the input and backend languages: Python
has \texttt{min}/\texttt{max} operators but C/C++ does not. % C++11 has...
C/C++ have
\texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}
% FIXME reviewer 4 doesn't seem to agree with and/or C++ vs Python
The latest OpenMP specifications~\cite{openmp4} describe a way to declare
user-defined reduction but is not implemented in any compiler yet.

@@

\begin{lstlisting}[float, label={lst:complex-omp-for}, caption={Parallel
    loop in Pythran with tuple unpacking.}]
'omp parallel for'
for i, v in enumerate([2, 3, 5, 7, 11]):
    print i, ':', v
\end{lstlisting}

@@

\begin{lstlisting}[float,label={lst:openmp-validation},caption={Example of
  Python OpenMP validation test case.}]
def omp_parallel_for_if(loop_count):
  import omp
  using = sum = 0
  'omp parallel for if(using == 1)'
  for i in range(loop_count + 1):
    num_threads = omp.get_num_threads()
    sum += i
  known_sum = (loop_count *
              (loop_count + 1)) / 2
  return known_sum == sum and num_threads == 1
\end{lstlisting}

@@

We then compare Pythran with Cython. These approaches share some similarities:
they both generate code written in a lower level language, with OpenMP
directives. However, the inputs differ as Cython requires more typing
information than Pythran to generate efficient code, and Cython is not
backward-compatible with Python. They translate explicit fine-grained
parallelism through code annotations or specific Python constructs,
respectively.

@@

    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython implementation
    of a parallel reduction.}]
from libc.math cimport sqrt
from cython.parallel import parallel, prange
def sum_sqrt(double r):
    cdef int i
    with nogil, parallel():
        for i in prange(10000000):
            r += sqrt(i)
    return r
    \end{lstlisting}
%
    \begin{lstlisting}[label={lst:pythran-sample}, caption={Pythran
      implementation of a parallel reduction.}]
#pythran export sum_sqrt(float)
import math
def sum_sqrt(r):
    #omp parallel for reduction(+:r)
    for i in xrange(10000000):
        r += math.sqrt(i)
    return r
    \end{lstlisting}
\end{figure}

@@

The Python Benchmark test bed
(\url{https://github.com/numfocus/python-benchmarks}) is an innitiative to
gather relevant numerical python benchmarks and compare several implementations
and compilers. The original benchmarks are slightly adapted to match each
compiler restriction (e.g. lack of support for a particular construct), or
completely rewritten in the target's paradigm (e.g. using PyOpenCL or Theano).
We picked up the targets that did not imply a full rewrite of their input:
Cython, Numba, Pythran and Parakeet, when available and supported.
Additionnaly, the original Python code is run.

The best execution time out of five run is shown for each code version. If the
code does not compile with a particular compiler, the column is left blank. Two
code versions are shown for Pythran: one compiler without the OpenMP flag set
and one with the OpenMP flag set. The code itself is not modified, only the
compilation flag. A logarithmic scale is used due to the generally poor
performance of the pure Python version.

@@

Figure~\ref{fig:pb-rosen} shows the execution time of the computation of the
third derivative of the Rosenbrock function for a random array input, using
Numpy's vector operations. There is no explicit parallelism in this function but
the vector operation is implictly parallel. On this example, Cython is on par
with Pythran and slightly faster than Parakeet, but turning OpenMP on makes the
Pythran version run $\times 1.78$ faster than sequential Pythran.

@@

Figure~\ref{fig:pb-growcut} shows the execution time of the computation of one
step of the GrowCut algorithm used in image processing. On this example,
Parakeet performs better than Cython and Pytrhan and way better than Numba, but
paralleizing the outermost loop with OpenMP gives a $\times 2.65$ speedup to
Pythran that helps it outperform Parakeet.

@@

It appears that all Python compilers manage to yield significant speedups over
the Python interepreter, each compiler managing some cases better than others.
All the benchmarks exhibt some level of parallelism that are easy to capture
using OpenMP without changing the original algorithm and while maintaining
backward compatibility with the Python interpreter. Some even have a degree of
implicit parallism that makes parallelization fully transparent to the user.
Taking advantage of this parallelism appears to be the key to reach an
additional performance level.

>>>>

Another widely used solution is the module \texttt{IPython.parallel}, which supports many
different styles of parallelism including: single program, multiple data (SPMD)
parallelism, multiple program, multiple data (MPMD) parallelism, message passing
using MPI, task farming, and data parallel.
%FIXME: to be continued...

@@

Figure~\ref{fig:pythran-compiler} summarizes its processing: type inference
helps removing most dynamic behavior, and is combined with a high-level runtime
library ---\texttt{Pythonic++}--- to allow a one-to-one mapping between Python
and C++ constructs. Pythonic++ is a template library, i.e.\ it consists
exclusively of a set of headers. Therefore the generated C++ file is easily
distributable and does not depend on Pythran. One could even preprocess this
C++ file to avoid the need of distributing along the Pythonic++ headers.

@@

At the AST level, it means that the Python AST is first reduced to a tree where
all nodes not available in C++ are transformed into a compatible representation.
For instance, list comprehension expressions are turned into function calls or
tuple unpacking is turned in multiple assignments enclosed in a dummy
\texttt{if}. During this transformation
process, an expression is always transformed into a single expression and a
statement is always transformed into a single statement.
Then this AST is converted into a C++ AST meant to be pretty-printed.

@@

The traditional solution is to use a native type that exposes a Python interface
and has a constant translation cost. NumPy's \texttt{ndarray} is a typical
implementation of this concept and is the keystone of the Scipy and NumPy
packages. Basically, a NumPy \texttt{ndarray} is a raw C pointer that is exposed
at the Python level. The PEP 3118 defined the API that allows to share
efficiently the embedded data between Python and the native world without
involving any copy. As a tool for scientific computing, Pythran supports such a
structure, implemented as a class wrapping a raw pointer.

@@

The \texttt{reduction(\emph{operator}: \emph{list})} directive is used to
characterize some data dependencies when performing a parallel reduction. The
list of supported operators depend on the input and backend languages: Python
has \texttt{min}/\texttt{max} operators but C/C++ does not. % C++11 has...
C/C++ have
\texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
the \texttt{and} and \texttt{or} keywords are not boolean operators in Python.}
The latest OpenMP specifications~\cite{openmp4} describe a way to declare
user-defined reduction but is not implemented in any compiler yet.

@@

\begin{lstlisting}[float, label={lst:complex-omp-for}, caption={Parallel
    loop in Pythran with tuple unpacking.}]
#omp parallel for
for i, v in enumerate([2, 3, 5, 7, 11]):
    print i, ':', v
\end{lstlisting}

@@

\begin{lstlisting}[float,label={lst:openmp-validation},caption={Example of
  Python OpenMP validation test case.}]
def omp_parallel_for_if(loop_count):
  import omp
  using = sum = 0
  #omp parallel for if(using == 1)
  for i in range(loop_count + 1):
    num_threads = omp.get_num_threads()
    sum += i
  known_sum = (loop_count *
              (loop_count + 1)) / 2
  return known_sum == sum and num_threads == 1
\end{lstlisting}

@@

We then compare Pythran with Cython. These approaches share some similarities:
they both generate code written in a lower level language, with OpenMP
directives. However, the inputs differ as Cython requires more typing
information than Pythran to generate efficient code, and Cython is not
backward-compatible with Python (outside of \emph{Pure mode}). They translate
explicit fine-grained parallelism through code annotations or specific Python
constructs, respectively.

@@

    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython implementation
    of a parallel reduction.}]
from libc.math cimport sqrt
from cython.parallel import parallel, prange
def sum_sqrt(double r):
    cdef int i
    for i in prange(10000000, nogil=True):
        r += sqrt(i)
    return r
    \end{lstlisting}
%
    \begin{lstlisting}[label={lst:pythran-sample}, caption={Pythran
      implementation of a parallel reduction.}]
#pythran export sum_sqrt(float)
import math
def sum_sqrt(r):
    #omp parallel for reduction(+:r)
    for i in xrange(10000000):
        r += math.sqrt(i)
    return r
    \end{lstlisting}
\end{figure}

@@

The Python Benchmark test bed
(\url{https://github.com/numfocus/python-benchmarks}) is an initiative to
gather relevant numerical python benchmarks and compare several implementations
and compilers. The original benchmarks are slightly adapted to match each
compiler restriction (e.g. lack of support for a particular construct), or
completely rewritten in the target's paradigm (e.g. using PyOpenCL or Theano).
We picked up the targets that did not imply a full rewrite of their input:
Cython, Numba, Pythran and Parakeet, when available and supported.
Additionally, the original Python code is run.

The best execution time out of five run is shown for each code version. If the
code does not compile with a particular compiler, the column is left blank. Two
code versions are shown for Pythran: one compiler without the OpenMP flag set
and one with the OpenMP flag set. The code itself is not modified, only the
compilation flag and the directives. A logarithmic scale is used due to the
generally poor performance of the pure Python version.

@@

Figure~\ref{fig:pb-rosen} shows the execution time of the computation of the
third derivative of the Rosenbrock function for a random array input, using
Numpy's vector operations. There is no explicit parallelism in this function but
the vector operation is implicitly parallel. On this example, Cython is on par
with Pythran and slightly faster than Parakeet, but turning OpenMP on makes the
Pythran version run $\times 1.78$ faster than sequential Pythran.

@@

Figure~\ref{fig:pb-growcut} shows the execution time of the computation of one
step of the GrowCut algorithm used in image processing. On this example,
Parakeet performs better than Cython and Pythran and way better than Numba, but
parallelizing the outermost loop with OpenMP gives a $\times 2.65$ speedup to
Pythran that helps it outperform Parakeet.

@@

It appears that all Python compilers manage to yield significant speedups over
the Python interpreter, each compiler managing some cases better than others.
All the benchmarks exhibit some level of parallelism that are easy to capture
using OpenMP without changing the original algorithm and while maintaining
backward compatibility with the Python interpreter. Some even have a degree of
implicit parallelism that makes parallelization fully transparent to the user.
Taking advantage of this parallelism appears to be the key to reach an
additional performance level.
